A Privacy Aware Mobile Sensor Application

### 2015-02-17 paper.tex
further changings from RG

\usepackage{amsmath}

@@

\begin{abstract}
In this article we analyze the privacy aspects of a mobile sensor
application used for recording urban travel patterns as part of a
travel-survey service. This service has been developed and
field-tested within the Live+Gov EU Project. The privacy analysis
follows a structured approach established in
\cite{Grimm:ItSecRefModel}. Eight privacy recommendations are derived,
and have already led to corresponding enhancements of the
travel-survey service.

@@

The rise of mobile smart-phones equipped with a wide range of sensors
and the abundance mobile internet connectivity has paved the way to a
whole new generation of mobile services which ease the daily life of
citizens using them. Apps like Google
Maps\footnote{\url{http://www.google.com/mobile/maps/}} allows the user to
navigate effectively in unknown cities; others track sports activities
in order to optimize training plans or engage in virtual competitions.

Besides the immediate benefits these services reveal a wealth of
private information about the citizen that capture a very detailed
picture of his life. For instance, GPS location tracking can be used
to infer shopping habits or associations with political groups (when
meetings are attended) and accelerometer data can be used to detect
medical conditions like walking disabilities. All this data is highly
sensitive to the citizens privacy and can be used against the citizen
if it falls into the wrong hands.

Providers of these new services are faced with a fundamental trade-off
between features and convenience of the service and the protection of
the citizen's privacy. The service has to collect enough data to
support the basic service promises. If too much data is collected the
privacy of the citizen is put at risk, with immediate implications for
the customer trust relation and the acceptance of the service. Also
leaks of private data can jeopardize the whole business.

@@

Travel surveys are regularly conducted by transport authorities which provide
public transportation infrastructure. With these surveys the current
usage of the public transportation system is assessed and the insights
are used for future planning of bus, tram and subway lines, etc..  For
each survey a large number of citizens (e.g. 5000) is asked to keep
track of their travel patterns for a constraint period of time
(e.g. one month). Traditionally the travel patterns have been recorded
manually in the form of travel diaries. Modern smart-phones are
equipped with a wide range of sensors (in particular GPS) that allow
the recording of travel-patterns in a fully automated way.

@@

Within the Live+Gov EU project, we have established a first
prototypical implementation of such an automated travel-survey
service. In this section we give a high level description of the
generic service architecture. This description along with the
introduced terminology will serve as a basis for our privacy analysis.

There are three stakeholders involved in the travel-survey service:
The \emph{citizen}, the \emph{service provider} and the \emph{transport
  authority}. The service provider is the operator of all IT systems and
offers the travel-survey service to the transport authority. Figure
\ref{figure:Live+Gov Vulnerabilities} contains a schematic
visualization of the relation between these stakeholders.

@@

The \emph{data center} stores and processes sensor data collected with
the sensor collector application. It can also take into account data
obtained from third parties, like the current positions of trains. In
particular, the data center determines if the citizen is using public
transportation and if so which lines of service (bus, train, tram) are
used. The data center can also sent mining end products and messages
(e.g. traffic jam reports, bus schedule) back to the mobile device of
the citizen.

The \emph{service provider} provides and operates all technical
infrastructure like the data center and the sensor application. He
also operates a web-based \emph{reporting tool} that aggregates
information about travel patterns of citizens within the
travel-survey. The reporting tool is used by the transport authority to gain
insights for planning and accounting of the public transportation
infrastructure.

@@

Modern mobile devices have a broad collection of sensors.
Disclosure or processing of sensor data can impact one's privacy.
In this section we identify groups of sensors and their potential impact on a certain type of privacy.
Figure \ref{figure:Sensor Data Privacy Impact Matrix} qualifies that impact on a simple scale.
Privacy of Data and Image is trivially threatened because here sensor data is individual data, a priori.
Indirect impact is caused by combining sensor data with additional knowledge.
For instance, comparing a contemporary map with locational data can imply behaviour if the position matches a church.

@@

\subsubsection{Motion Sensors.}
Accelerometer, Rotation Vector, Gyroscope and Magnetic field sensor measure the physical movement of the mobile device on all three axes.
If the mobile device is carried ``normally'' its safe to say that those sensors also measure the moments of its carrier.
So his privacy is infringed regarding biometric behaviour, i.e. the Privacy of the Person.

@@

\subsubsection{Network Sensors.}
The GSM and WLAN sensors reveal the position of the mobile device and its carrier, when used in connection with external databases.
The both sensors give the exact cell or network, the mobile device has registered with at the current moment.
Frequent connection to one particular network also reveals association, e.g. university networks. The Bluetooth sensors record lists of the bluetooth clients in the direct neighbourhood.
Since those clients are usually moving, inference of the position is usually not possible.
Instead, bluetooth clients carried by a third person may infringe the Privacy of Association.

@@

In sections \ref{subsec:world-analysis} and \ref{subsec:pot-analysis} we apply the first two steps of the reference model \cite{Grimm:ItSecRefModel}, which are related to the views on the real world and on the measures and plans. In section \ref{subsec:priv-rec} we apply the third step of the reference model by providing specific privacy recommendations and requirements as a result of the previous analysis that the system must comply with.

@@

The relevant \textbf{IT-Systems} were already described in Section \ref{sec:intro}.
\textbf{Interactions} between assets, IT-Systems, humans, and vulnerabilites as partly described in Section \ref{sec:intro} and will be further analyzed in Section \ref{subsec:pot-analysis}.

@@

In this scenario we focus our attention to only one asset: The privacy of the citizen. Our definition of privacy is described in detail in Section \ref{sec:privacydef}.

@@

This section describes the human actors previously introduced in section \ref{sec:ScenarioDescription}. Although the text is only a short description, a list of the most important interests is provided for each actor.

\textbf{Citizen.}
Citizens are persons who use the mobile device as users of the provided software.
Their main motivation for using the software is to gain a higher level of convenience in their daily activities. For example they may have access to real time bus schedules or reports about traffic jams, that help them to avoid long waiting times. Also they generally benefit from improvements of public infrastructure by travel agencies, which is triggered by issue reports. By using the application, citizens are sharing personal information like name and address, as well as data gathered from mobile sensor with the service provider. This data can be exploited in ways that are harmful to the citizen (cf. \cite{GuardienMassSurveillance}). Citizens are interested in: physical wellbeing and health, financial profit, convenience, legitimate use of personal data, non-disclosure of personal data to peers of the citizen, and not being monitored.

\textbf{External.}
Externals are persons who do not have privileged access to the IT systems, and are willing to break laws, security constrains and norms in order to promote their interests. A common interest of an external is financial profit. For example they want to obtain access to critical systems to steal sensitive data or to get the system under their control. Stolen data could simply be sold as is or used for illegitimate purposes, e.g. spam or phishing attacks - or excessive data mining. In short, externals could be interested in: increase power over citizen, financial profit, political activism, and their social standing.

\textbf{System Provider.}
System Providers operate the technical infrastructure (hardware and software) of the IT System. They are private companies and legal persons in their own right, but also employ a number of people with diverging interests including administrators, programmer/developer, and a support manager. As companies, they are interested in gaining financial profit. The financial success of system providers depends on the task complexity of the maintained infrastructure. The complexity of a task has to be in reasonable bounds, so that system providers can complete it within time, with a satisfying quality. System Providers are interested in: financial profit, manageable complexity, professional excellence, and good working conditions.

\textbf{Transport Authority.}
 Transport authorities are public offices (ministry, agency, department, ...) or other external public entities which act as direct customers of service providers. For example a department for urban mobility, orders a system to better understand usage patterns and make improvement to the urban traffic flow. Such systems are investments, and so naturally travel agencies are interested in a profitable return, like increased ticket sales. Transport authorities are interested in: financial profit, political reputation, business intelligence, and good working conditions.

\subsubsection{Conflicts of Interests}
\label{subsubsection:Conflicts of Interests}
Different actors have different interests which can be in conflict each other. This section outlines the Conflicts of Interests between the actors of the proposed IT system architecture. The individual interests of all actors is already described in the previous section and are not elaborated any further. The emphasis here is put on prominent existing conflicts, because they provide a foundation for vulnerabilities and subsequent threats.

@@

\textbf{Power of External vs. Privacy.}
Externals which are in a social relation to the citizen can have an interest in obtaining further information in order to gain power. In the most simplistic example this could be a man wanting monitor the activities of his spouse.

\textbf{Financial Profit of External vs Privacy.}
Externals can gain financial profit from stealing privacy sensitive data.
For example by selling raw contact information to advertisers or by selling mined data to insurance companies, or intermediaries like scoring companies. In such cases, citizens lose complete control over their data.

@@

\subsubsection{Vulnerabilities}
\label{subsubsection:Vulnerabilities}
This section outlines the vulnerabilities (Figure \ref{figure:Live+Gov Vulnerabilities}) of the proposed monitoring and mining system.

@@

\textbf{Unhappy Employees.}
An Employee that is frustrated with his situation for a long time
period constitutes a security vulnerability. On the one hand he might
want to harm his employer directly, on the other he is increasingly
susceptible for social engineering.

\textbf{Inadequate Access Rules.}
The proposed IT system infrastructure has various accesses to privacy sensitive data.
System Provider staff has access to Data Center hardware and software like databases, web-servers and other inspection tools. Transport authority staff has access to the Report Tool. This all enables staff members to have potential access to privacy sensitive information.

@@

The \emph{Potential World Analysis} displays the intended and unintended interactions of the components in the world view. The intended interactions support the underlying business objectives. Unintended interactions can lead to threats. Obviously this provides a conflict of interest with the victim's business model, to keep the asset away from unauthorized access. From the point of view of the victim, an attack would be an unintended interaction. The potential world view consists of the following components: Business Objectives, Threats, hances/Risk, and Security Requirements. 
\textbf{Business Objectives} of the system owner were already described in Section \ref{sec:intro} and \ref{subsec:world-analysis}. \textbf{Security Requirements} are included to our analysis in Section \ref{subsec:priv-rec}.

@@

A threat is a potential interaction of the components that targets an asset.
We restrict ourselves to the case of attacks, and the asset of privacy.
An attack is an interaction that is executed by an actor in response to a conflict of interest by exploiting a vulnerability of the system. We describe threats for the citizen privacy in the system. This list can necessarily not be complete, but we make a best effort to cover the most relevant cases.

@@

\textbf{T2. Excessive Data Mining.}
The System Provider and/or the transport authority secretly extract more private information from the collected data, than the Citizen agreed to. But results of the mining process create no disadvantages for Citizens, because there is no disclosure to third parties. This could be the case for a System Provider, who wants to test a new product and uses the pre-existing data collection. Thus there are two possible conflicts: \textit{Privacy vs. Financial Profit of Serivce Provider} and  \textit{Privacy vs. Buisness Intelligence of Transport Authority} The threat can be provoked by either lax data handling policies of both System Providers and Transport Authorities, or a weak law enforcement of existing supervision. But the main issue, which can lead to such threats, is again a general \textit{Missing Privacy Awareness}.

\textbf{T3. Data Theft}
An External infiltrates infrastructure in order to steal personal data and sell it on the black market. Also the External might be motivated politically and wants to harm the reputation of the System Provider or the Transport Authority. Such a successful attack could harm the reputation of both System Provider and Transport Authority. Thus this threat is defined by three conflicts: \textit{Privacy vs. Financial Profit}, \textit{Reputation of Service Provider vs. Political Activism} and \textit{Reputation of Transport Authority vs. Political Activism}. Also this threat describes the classical scenario, where attacks are provoked by \textit{Insecure Infrastructure} (SQL injection) and \textit{Insecure Communication} (Packet Capture).

\textbf{T4. Surveillance}
An External infiltrates infrastructure in order to obtain information
about the citizen and exploit it directly.  In this scenario the
external is supposed to have some direct relationship to the citizen
which motivates his interest to obtain personal information.  Examples
could be a public institution that wants to gain information about
planned activities of the citizens (e.g. Nixon's Watergate scandal or
the recent prosecution of Guardian journalists by GHCQ). 
In this threat the privacy interest of the citizen is in conflict with
the aspirations for power over the citizen by the externals.

\textbf{T5. Information Leak} 
Like an external person the Data Theft Scenario an employee of the
service provider or the transport authority has selfish interests to gain
money, make political statements or harm his employer.  In order to
pursue this interest he can steal personal data and sell it or release
it to the public.  The corresponding conflicts of interests are:
\textit{Privacy vs. Financial Profit} of the Employee, \textit{Reputation of Service
  Provider vs. Political Activism} of the Employee and \textit{Reputation of Transport
  Authority vs. Political Activism} of the Employee.  The vulnerability constitutes of
the existence of \textit{Unhappy employees} itself and possibly \textit{lax access
rules} that enable the employee to obtain large amounts of data
unnoticed.

\textbf{T6. Social Engineering}
This scenario an external manipulates an employee of a Service
Provider or the Transport Authority to leak information to the external
person.  It is thus combination of the Data Theft and Information Leak
scenario.
The conflicts of interest are \textit{Privacy vs. Financial Profit} of
the External, \textit{Reputation of Service Provider vs. Political
  Activism} of the external and \textit{Reputation of Transport Authority
  vs. Political Activism} of the external.  The exploited
vulnerabilities are, again, the existence of \textit{Unhappy
  employees} and possibly \textit{lax access rules} that enable the
employee to obtain large amounts of data unnoticed.

@@

In this section we will associate to every identified threat a corresponding risk. A risk is the expected loss that is associated to the threat.
Therefore, we have to quantify the likeliness of the threat to occure and the harm or loss done in this case.
The quantification of likeliness will be solely based on rough judgment of the authors.
The quantification of loss, will be made in a two step process.
For each threat listed in the previous section, we have analyzed the affected personal data of the citizen.
For each possible data type (e.g. GPS) we analyze the impact on the seven different types of privacy in Section \ref{sec:SensorPrivacyImpact}.
In combination we can quantify roughly the impact of each threat on the citizens privacy. Both evaluations are necessarily fraught with a high level of uncertainty.

For the quantification of the loss in case of a threat scenario we use the following rough calibration between 3 (high) and 0 (none). For the quantification of likeliness the following scale between 4 (always) and 0 (impossible) is used. The quantification of the risk, we add the values for loss and likeliness of the corresponding threats. Note, that loss and
likeliness scales have a logarithmic character, so that addition
of those scales corresponds to multiplication of the usual scales. The likeliness, loss and the resulting risks assigned to the threats
are discussed in the following paragraphs and summarized in Figure
\ref{fig:risks}.

\textbf{Risk of T1. Insufficient Control Features.}  The occurrence of this
threat is dependent on the design on the system and given in our case,
since we do not give the citizen control over his data once it is
recorded. Therefore the Likeliness is evaluated as $4$ (Always).  The
associated, risk is $1$ Low on our scale, since no direct harm is done to
the citizen by exploiting the data. Hence the resulting risk is calculated as $4+1 = 5$.

\textbf{Risk of T2. Excessive Data Mining.}
We assess the likeliness of excessive data mining to be $3$ High, since
these kind of analysis can be performed within the walls of the service
provider, without somebody else noticing, and the service provider himself
has an interest in this activity. The associated loss, on the other hand can be substantial (Medium $2$). Hence the resulting risk is calculated as $3+2 = 5$.

\textbf{Risk of T3. Data Theft.}
The likeliness of a targeted attack by a third party is dependent on
the popularity of the offered service and financial value of the
captured information. Moreover, the amount of manual work required to
infiltrate a custom build system is significantly higher that that of
compromising a standard software solution. In the scenario we assume a
moderate popularity in a single metropolitan area, with around 10.000
users and storage of data of only limited financial value (no
addresses, no payment information). Therefore the likeliness
assessment is $1-2$ (Low-Medium). The harm of leaked information to a criminal party is $3$ High. Hence the resulting risk is calculated as $4-5$.

\textbf{Risk of T4. Surveillance.}  In the surveillance scenario an party
related to the citizen, like a company where he is customer of, or a
government agency, seeks to obtain sensitive information from our
service. The likeliness of such an intrusion is hard to asses, and depends
again on the popularity of the service. If a high popularity is
reached we have recently learned that spying by government agencies is
very likely to occur. The barrier for companies that do not operate
the infrastructure used to transmit the data a surveillance attack is
however very hard to perform. Therefore we assess the likeliness of
the threat with $1-2$ (Low-Medium). The harm of leaked information to a related party is $3$ High. Hence the resulting risk is calculated as $4-5$.

@@

In this section we derive recommendations or
requirements for a system that addresses these risks. The mapping of these recommendations to the threats is summarized in Figure \ref{fig:risks}.

\textbf{R1. Privacy Dashboard}. In order to address the threat with the highers risk, Insufficient Control (T1) of the citizen, we need to give the citizen back the
control over its data inside the system. The most direct way to do
this is to provide a web-based overview which allows the citizen to view, edit and delete all information about his person
that is stored inside the system.

@@

\textbf{R3. Data Handling Guidelines}. The threat with the second largest risk is (T2) Excessive Data Mining.
Contrary to common belief, it is neither legal nor ethical to process
personal data for by new methods or for new purposes that were not
stated and explained to the citizen at the time of data
collection. Also the common practice of obtaining far-reaching
permissions from the citizens inside the privacy policy is neither an
ethical or legal solution to the problem. To address this threat awareness about the limitations of data
processors inside the company is a key element. As one mean to
establish such a culture of privacy respect, we recommend to prepare
an guideline intended for internal use that explains the concrete processing steps and purposes that are permitted by the citizens.

\textbf{R4. User Permissions}. If further processing should be performed, it is necessary to seek additional permissions from the citizen. A simple email explaining the planned processing steps, and asking for permissions would be enough for this purpose. The permission can be given via an embedded link that shall be followed in order to signal agreement.

\textbf{R5. Anonymization}. An alternative measure to address the risk of excessive data mining is the anonymization of data. When all direct- or indirect
links to the identity of the person are removed, no violation of the
citizens privacy caused by arbitrary processing. However removing all
such links is a challenging tasks, and full anonymity is often not
achieved, cf. \cite{krumm2009}.

>>>>

%\usepackage{amsmath}

@@

\begin{abstract}
In this article we analyze the privacy aspects of a mobile sensor
application used for recording urban travel patterns as part of a
travel-survey service. This service has been developed and
field-tested within the Live+Gov EU Project \cite{LiveGov:Prog}. The privacy analysis
follows a structured approach established in
\cite{Grimm:ItSecRefModel}. Eight privacy recommendations are derived,
and have already led to corresponding enhancements of the
travel-survey service.

@@

The rise of mobile smart-phones equipped with a wide range of sensors
and the abundance mobile Internet connectivity has paved the way to a
whole new generation of mobile services which ease the daily life of
citizens using them. Apps like Google
Maps\footnote{\url{http://www.google.com/mobile/maps/}} allows the user to
navigate effectively in unknown cities; others track sports activities
in order to optimize training plans or engage in virtual competitions.

Besides the immediate benefits these services reveal a wealth of
private information about the citizen that capture a very detailed
picture of his life. For instance, GPS location tracking can be used
to infer shopping habits or associations with political groups (when
meetings are attended) and accelerometer data can be used to detect
medical conditions like walking disabilities. All this data is highly
sensitive to the citizen's privacy and can be used against the citizen 
if it falls into the wrong hands.

Providers of these new services are faced with a fundamental trade-off
between features and convenience of the service and the protection of
the citizen's privacy. The service has to collect enough data to
support the basic service promises. If too much data is collected the
privacy of the citizen is put at risk, with immediate implications for
the customer trust relation and the acceptance of the service. Also,
leaks of private data can jeopardize the whole business.

@@

Travel surveys are regularly conducted by transport authorities which provide
a public transportation infrastructure. With these surveys the current
usage of the public transportation system is assessed and the insights
are used for future planning of bus, tram and subway lines, etc.  For
each survey a large number of citizens (e.g. 5000) is asked to keep
track of their travel patterns for a constraint period of time
(e.g. one month). Traditionally the travel patterns have been recorded
manually in the form of travel diaries. Modern smart-phones are
equipped with a wide range of sensors (in particular GPS) that allow
the recording of travel-patterns in a fully automated way.

@@

Within the Live+Gov EU project \cite{LiveGov:Prog}, we have established a first
prototypical implementation of such an automated travel-survey
service. In this section we give a high level description of the
generic service architecture. This description along with the
introduced terminology will serve as a basis for our privacy analysis.

There are three stakeholders involved in the travel-survey service:
The \emph{citizen}, the \emph{service provider} and the \emph{transport
  authority}. The service provider is the operator of all IT systems and
offers the travel-survey service to the transport authority. Figure
\ref{figure:Live+Gov Vulnerabilities} in section \ref{subsubsection:Conflicts of Interests} below contains a schematic
visualization of the relation between these stakeholders.

@@

The \emph{data center} is under the control of the service provider. It 
stores and processes sensor data collected with
the sensor collector application. It can also take into account data
obtained from third parties, like the current positions of trains. In
particular, the data center determines if the citizen is using public
transportation and if so which lines of service (bus, train, tram) are
used. The data center can also send mining end products and messages
(e.g. traffic jam reports, bus schedule) back to the mobile device of
the citizen.

The \emph{service provider} provides and operates all technical
infrastructure like the data center and the sensor application. He
also operates a web-based \emph{reporting tool} that aggregates
information about travel patterns of citizens within the
travel-survey. The reporting tool is used by the \emph{transport authority} to gain
insights for planning and accounting of the public transportation
infrastructure.

@@

Modern mobile devices have a broad collection of sensors.
Disclosure or processing of sensor data can impact one's privacy.
In this section we identify groups of sensors and their potential impact on a certain type of privacy.
Figure \ref{figure:Sensor Data Privacy Impact Matrix} qualifies that impact on a simple scale.
Privacy of Data and Image is trivially threatened because here sensor data is individual data, a priori.
Indirect impact is caused by combining sensor data with additional knowledge.
For instance, comparing a contemporary map with locational data can reveal behaviour and thoughts if the position matches a church.

@@

\subsubsection{Motion Sensors.}
Accelerometer, Rotation Vector, Gyroscope and Magnetic field sensor measure the physical movement of the mobile device on all three axes.
If the mobile device is carried ``normally'' its safe to say that those sensors also measure the moments of its carrier.
So privacy is infringed regarding biometric behaviour, i.e. the Privacy of the Person.

@@

\subsubsection{Network Sensors.}
The GSM and WLAN sensors reveal the position of the mobile device and its carrier, when used in connection with external databases (location).
Both sensors give the exact cell or network, the mobile device has registered with at the current moment.
Frequent connection to one particular network also reveals Association, e.g. university networks. The Bluetooth sensors record lists of the bluetooth clients in the direct neighbourhood.
Since those clients are usually moving, inference of the position is usually not possible.
Instead, bluetooth clients carried by a third person may infringe the Privacy of Association.

@@

In sections \ref{subsec:world-analysis} and \ref{subsec:pot-analysis} we apply the first two steps of the reference model \cite{Grimm:ItSecRefModel}, which are related to the views on the real world and its privacy risks. In section \ref{subsec:priv-rec} we apply the third step of the reference model by providing specific recommendations and for privacy measures as a result of the previous analysis that the system must comply with.

@@

The relevant \textbf{IT-Systems} were already described in section \ref{sec:intro}.
\textbf{Interactions} between assets, IT-Systems, humans, and vulnerabilites is partly described in section \ref{sec:intro} and will be further analyzed in section \ref{subsec:pot-analysis}.

@@

In this scenario we focus our attention to only one asset: The privacy of the citizen. Our definition of privacy is described in detail in section \ref{sec:privacydef}.

@@

This section describes the human actors previously introduced in section \ref{sec:ScenarioDescription}. A short description for each actor includes a list of the most important interests.

\textbf{Citizen.}
Citizens are persons who use the mobile device as users of the provided software.
Their main motivation for using the software is to gain a higher level of convenience in their daily activities. For example they may have access to real time bus schedules or reports about traffic jams that help them to avoid long waiting times. Also they generally benefit from improvements of public infrastructure by transport authorities, which is triggered by issue reports. By using the application, citizens are sharing personal information like name and address, as well as data gathered from mobile sensors with the service provider. This data can be exploited in ways that are harmful to the citizen (cf. \cite{GuardienMassSurveillance}). Citizens are interested in: physical wellbeing and health, financial profit, convenience, legitimate use of personal data, non-disclosure of personal data to peers of the citizen, and not being monitored.

\textbf{Malicious External.}
Malicious externals are persons who do not have privileged access to the IT systems, and are willing to break laws, security constrains and norms in order to promote their interests. A common interest of an external is financial profit. For example they want to obtain access to critical systems to steal sensitive data or to get the system under their control. Stolen data could simply be sold as is or used for illegitimate purposes, e.g. spam or phishing attacks - or excessive data mining. In short, externals could be interested in: increase power over citizen, financial profit, political activism, and their social standing.

\textbf{System Provider.}
System Providers operate the technical infrastructure (hardware and software) of the IT System. They are private companies and legal persons in their own right, but also employ a number of people with diverging interests including administrators, programmer/developer, and a support manager. As companies, they are interested in gaining financial profit. The financial success of system providers depends on the task complexity of the maintained infrastructure. The complexity of a task has to be in reasonable bounds, so that system providers can complete it within time, with a satisfying quality. System Providers are interested in: financial profit, manageable complexity, professional excellence, and good working conditions, and a good reputation.

\textbf{Transport Authority.}
 Transport authorities are public offices (ministry, agency, department, ...) or other external public entities which act as direct customers of service providers. For example a department for urban mobility orders a system to better understand usage patterns and make improvement to the urban traffic flow. Such systems are investments, and so naturally transport authorities are interested in a profitable return, like increased ticket sales. Transport authorities are interested in: financial profit, political reputation, business intelligence, and good working conditions.

\subsubsection{Conflicts of Interests}
\label{subsubsection:Conflicts of Interests}
Different actors have different interests which can be in conflict with each other. This section outlines the Conflicts of Interests between the actors of the proposed IT system architecture. The emphasis here is to put on prominent existing conflicts, because they provide a foundation for vulnerabilities and subsequent threats.

@@

\textbf{Power of External vs. Privacy.}
Malicious externals which are in a social relation to the citizen can have an interest in obtaining further information in order to gain power. In the most simplistic example this could be a man wanting to monitor the activities of his spouse.

\textbf{Financial Profit of External vs Privacy.}
Malicious Externals can gain financial profit from stealing privacy sensitive data.
For example by selling raw contact information to advertisers or by selling mined data to insurance companies, or intermediaries like scoring companies. In such cases, citizens lose complete control over their data.

@@

\subsubsection{Vulnerabilities}
\label{subsubsection:Vulnerabilities}
This section outlines the vulnerabilities of the proposed monitoring and mining system. This is illustrated by figure \ref{figure:Live+Gov Vulnerabilities} below.

@@

\textbf{Unhappy Employees.}
As an insider, an Employee who is frustrated with his situation constitutes a security vulnerability. On the one hand he might
want to harm his employer directly, on the other he is increasingly
susceptible for social engineering.

\textbf{Inadequate Access Rules.}
The proposed IT system infrastructure has various accesses to privacy sensitive data.
System Provider staff has access to Data Center hardware and software like databases, web-servers and other inspection tools. Transport authority staff has access to the Report Tool. Moreover, access control systems are security bottlenecks,

@@

The \emph{Potential World Analysis} displays the intended and unintended interactions of the components in the world view. The intended interactions support the underlying business objectives. Unintended interactions can lead to threats. Obviously this provides a conflict of interest with the victim's business model, to keep the asset away from unauthorized access. From the point of view of the victim, an attack would be an unintended interaction. The potential world view consists of the following components: Business Objectives, Threats, Chances/Risks, and Security Requirements. 
\textbf{Business Objectives} of the system owner were already described in section \ref{sec:intro} and \ref{subsec:world-analysis}. \textbf{Security Requirements} are included to our analysis in section \ref{subsec:priv-rec}.

@@

A threat is a potential interaction of the components that targets an asset.
We restrict ourselves to the case of attacks, and the asset of privacy.
An attack is an interaction that is executed by an actor in response to a conflict of interest by exploiting a vulnerability of the system. We describe threats for the citizen's privacy in the system. As a matter of course, this list cannot be complete, but we make a best effort to cover the most relevant cases.

@@

\textbf{T2. Excessive Data Mining.}
The System Provider and/or the transport authority secretly extract more private information from the collected data, than the Citizen agreed to. But results of the mining process create no disadvantages for Citizens, because there is no disclosure to third parties. This could be the case for a System Provider, who wants to test a new product and uses the pre-existing data collection. Thus there are two possible conflicts: \textit{Privacy vs. Financial Profit of Serivce Provider} and  \textit{Privacy vs. Buisness Intelligence of Transport Authority}. The threat can be provoked by either careless data handling policies of both System Providers and Transport Authorities, or a weak law enforcement of existing supervision. But the main issue, which can lead to such threats, is again a general \textit{Missing Privacy Awareness}.

\textbf{T3. Data Theft}
An malicious external infiltrates the infrastructure in order to steal personal data and sell it on the black market. Also the External might be motivated politically and wants to harm the reputation of the System Provider or the Transport Authority. Such a successful attack could harm the reputation of both System Provider and Transport Authority. This threat is defined by three conflicts: \textit{Privacy vs. Financial Profit}, \textit{Reputation of Service Provider vs. Political Activism} and \textit{Reputation of Transport Authority vs. Political Activism}. Also this threat describes the classical scenario, where attacks are provoked by \textit{Insecure Infrastructure} (SQL injection) and \textit{Insecure Communication} (Packet Capture).

\textbf{T4. Surveillance}
An External infiltrates infrastructure in order to obtain information
about citizens and exploit it directly.  In this scenario the
external is supposed to have some direct relationship to the citizen
which motivates his interest to obtain personal information.  Examples
could be a public institution that wants to gain information about
planned activities of the citizens (e.g. Nixon's Watergate scandal or
the recent prosecution of Guardian journalists by GHCQ). 
In this threat the privacy interest of the citizen is in conflict with
the aspirations for power over citizens by the externals.

\textbf{T5. Information Leak} 
Like a malicious external person in the Data Theft scenario an employee of the
service provider or the transport authority has selfish interests to gain
money, make political statements or harm his employer.  In order to
pursue this interest he can steal personal data and sell it or release
it to the public.  The corresponding conflicts of interests are:
\textit{Privacy vs. Financial Profit} of the Employee, \textit{Reputation of Service
  Provider vs. Political Activism} of the Employee and \textit{Reputation of Transport
  Authority vs. Political Activism} of the Employee.  The vulnerability constitutes of
the existence of \textit{Unhappy employees} and possibly \textit{careless access
rules}, which enable the employee to obtain large amounts of data
unnoticed.

\textbf{T6. Social Engineering}
In this scenario an external manipulates an employee of a Service
Provider or the Transport Authority to leak information to the external
person.  It is thus a combination of the Data Theft and Information Leak
scenario.
The conflicts of interest are \textit{Privacy vs. Financial Profit} of
the external, \textit{Reputation of Service Provider vs. Political
  Activism} of the external and \textit{Reputation of Transport Authority
  vs. Political Activism} of the external.  The exploited
vulnerabilities are, again, the existence of \textit{Unhappy
  employees} and possibly \textit{careless access rules} that enable the
employee to obtain large amounts of data unnoticed.

@@

In this section we will associate to every identified threat with a corresponding risk. A risk is the expected loss that is associated to the threat.
Therefore, we have to quantify the likeliness of the threat to occure and the harm or loss done in this case.
The quantification of likeliness will be solely based on rough judgment of the authors.
The quantification of loss will be made in a two step process.
For each threat listed in the previous section, we have analyzed the affected personal data of the citizen.
For each possible data type (e.g. GPS) we analyze the impact on the seven different types of privacy in section \ref{sec:SensorPrivacyImpact}.
In combination we can quantify roughly the impact of each threat on the citizens privacy.

For the quantification of the loss in case of a threat scenario we use the following rough calibration between 3 (high) and 0 (none). For the quantification of likeliness the following scale between 4 (always) and 0 (impossible) is used. For the quantification of the risk, we add the values for loss and likeliness of the corresponding threats. Note, that loss and
likeliness scales have a logarithmic character, so that addition
of those scales corresponds to multiplication of the usual scales. The likeliness, loss and the resulting risks assigned to the threats
are discussed in the following paragraphs and summarized in figure
\ref{fig:risks}.

\textbf{Risk of T1. Insufficient Control Features.}  This threat depends on the design of the system. In our case it is always there,
since we do not give the citizen control over his data once it is
recorded. Therefore the Likeliness is evaluated as $4$ (Always).  The
associated, risk is $1$ Low on our scale, since no direct harm is done to
the citizen by exploiting the data. Hence the resulting risk is calculated as $4+1 = 5$.

\textbf{Risk of T2. Excessive Data Mining.}
We assess the likeliness of excessive data mining to be $3$ High, since
this kind of analysis can be performed within the walls of the service
provider, without somebody else noticing, and the service provider himself
has an interest in this activity. The associated loss, on the other hand can be substantial (Medium $2$). Hence the resulting risk is calculated as $3+2 = 5$.

\textbf{Risk of T3. Data Theft.}
The likeliness of a targeted attack by a third party dependens on
the popularity of the offered service and on the financial value of the
captured information. Moreover, the amount of manual work required to
infiltrate a custom build system is significantly higher that that of
compromising a standard software solution. In the scenario we assume a
moderate popularity in a single metropolitan area, with around 10.000
users and storage of data of only limited financial value (no
addresses, no payment information). Therefore the likeliness
assessment is $1-2$ (Low-Medium). The harm of leaked information to a criminal party is $3$ High. Hence the resulting risk is calculated as $4-5$.

\textbf{Risk of T4. Surveillance.}  In the surveillance scenario a party
related to the citizen, like a company that he is customer of, or a
government agency, seeks to obtain sensitive information from our
service. The likeliness of such an intrusion is hard to asses, and depends
again on the popularity of the service. If a high popularity is
reached we have recently learned that spying by government agencies is
very likely to occur. The barrier for companies that do not operate
the infrastructure used to transmit the data a surveillance attack is very 
hard to break. Therefore we assess the likeliness of
the threat with $1-2$ (Low-Medium). The harm of leaked information to a related party is $3$ High. Hence the resulting risk is calculated as $4-5$.

@@

In this section we derive recommendations or
requirements for a system that addresses these risks. The mapping of these recommendations to the threats is summarized in figure \ref{fig:risks}.

\textbf{R1. Privacy Dashboard}. In order to address the threat with the highest risk, Insufficient Control (T1) of the citizen, we need to give the citizen back the
control over his data inside the system. The most direct way to do
this is to provide a web-based overview which allows the citizen to view, edit and delete all information about his person
that is stored inside the system.

@@

\textbf{R3. Data Handling Guidelines}. The threat with the second largest risk is (T2) Excessive Data Mining.
Contrary to common belief, it is neither legal nor ethical to process
personal data by new methods or for new purposes that were not
stated and explained to the citizen at the time of data
collection. Also the common practice of obtaining far-reaching
permissions from the citizens inside the privacy policy is neither an
ethical or legal solution to the problem. To address this threat awareness about the limitations of data
processors inside the company is a key element. As one mean to
establish such a culture of privacy respect, we recommend to prepare
an guideline intended for internal use that explains the concrete processing steps and purposes that have been permitted by the citizens.

\textbf{R4. User Permissions}. If further processing should be performed, it is necessary to seek additional permissions from the citizens. A simple email explaining the planned processing steps, and asking for permissions would be enough for this purpose. The permission can be given via an embedded link that shall be followed in order to signal agreement.

\textbf{R5. Anonymization}. An alternative measure to address the risk of excessive data mining is the anonymization of data. When all direct- or indirect
links to the identity of the person are removed, no violation of the
citizen's privacy is caused by arbitrary processing. However, removing all
such links is a challenging tasks, and full anonymity is often not
achieved, cf. \cite{krumm2009}.
